---
title: 'MA304 Data Visualization'
author: "Assignment MA-304 "
date: "Melek Kuru (mk23930@essex.ac.uk) "
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, results = "markup")
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("dplyr")
install.packages("ggplot2",dependencies=TRUE)
install.packages("viridisLite")
library(dplyr)
library(zoo)
library(ggplot2)
library(ggthemes)
library(tidyr)
library(ggcorrplot)
library(plotly)
library(xts)
library(lubridate)
install.packages("rmarkdown")
library(dplyr)
library(knitr)
install.packages("webshot")
webshot::install_phantomjs()  # Only necessary if using `webshot`

# Or for webshot2
install.packages("webshot2")
knitr::opts_chunk$set(webshot = "webshot2")  # If using `webshot2`

```



```{r, eval=TRUE}
# Read the temperature file and assign to a more descriptive variable
temperature_data <- read.csv("temp2023.csv", header = TRUE)
str(temperature_data)

# Read the crime file and assign to a more descriptive variable
crime_records <- read.csv("crime23.csv", header = TRUE)
str(crime_records)

```
**DATA PREPROCESSING:**

We prepared our Colchester datasets for analysis by ensuring that the tools
`dplyr` and `xts` worked well together. We filled in missing data points in
the temperature records and standardized the dates in the crime records.
With these datasets pruned and primed, we’re ready to dive into
the heart of Colchester’s climate and crime stories, armed with 
data that’s now as clear as a crisp spring morning.
```{r, eval=TRUE}
# Configure dplyr to avoid conflicts and set options for xts
conflictRules('dplyr', exclude = 'lag')
options(xts.warn_dplyr_breaks_lag = FALSE)

# Check if there are missing values in the temperature data variables
check_miss_values <- dplyr::summarise_at(temperature_data, vars(everything()),
                            list(n = ~sum(!is.na(.)), missing = ~sum(is.na(.))))
check_miss_values

# Drop SnowDepcm & PreselevHp variables then check
#the dimensions/names of the cleaned data
temperature_cleaned = select(temperature_data, -c(SnowDepcm, PreselevHp))
dim(temperature_cleaned)

# Check for duplicate rows in the cleaned temperature data
temperature_cleaned = distinct(temperature_cleaned)
dim(temperature_cleaned)


# Replace missing values in the Precmm column with zeros
# Assuming most of the missing values are in summer
temperature_cleaned$Precmm[is.na(temperature_cleaned$Precmm)] <- 0

# Replace missing values in the lowClOct column with the median value
median_lowClOct <- median(temperature_cleaned$lowClOct, na.rm = TRUE)
temperature_cleaned$lowClOct[is.na(temperature_cleaned$lowClOct)] <- median_lowClOct

# Replace missing values in the SunD1h column with the median value
median_SunD1h <- median(temperature_cleaned$SunD1h, na.rm = TRUE)
temperature_cleaned$SunD1h[is.na(temperature_cleaned$SunD1h)] <- median_SunD1h

# Rename the data subset for clarity
temperature_final <- temperature_cleaned

# Check if there are missing values in the variables after handling them
missing_values_final <- sum(is.na(temperature_final))
cat("Number of missing values after imputation:", missing_values_final)

# Change the date character to date format using lubridate's ymd function
tsdates <- ymd(temperature_final$Date)
temperature_final$Date <- tsdates

# Check the names of the variables in the final temperature dataset
names(temperature_final)


```

```{r, eval=TRUE}
# Check if there are missing values in the crime records dataset
check_miss_values_crime <- dplyr::summarise_all(crime_records, 
                          list(n = ~sum(!is.na(.)), missing = ~sum(is.na(.))))
check_miss_values_crime

# Drop location_subtype & context variables then check the 
#dimensions/names of the cleaned data
crime_records_cleaned = select(crime_records, -c(context, location_subtype))
dim(crime_records_cleaned)

# Check for duplicate rows in the cleaned crime data
crime_records_cleaned = distinct(crime_records_cleaned)
dim(crime_records_cleaned)

# Change the date character to date format by appending "-01"
#to make it a full date string
tsdates_crime <- paste0(crime_records_cleaned$date, "-01")
crime_records_cleaned$date <- as.Date(tsdates_crime, format = "%Y-%m-%d")

# Rename the data subset for clarity
crimes_final <- crime_records_cleaned

# Check the names of the variables in the final crime dataset
names(crimes_final)



```
```{r, eval=TRUE}
# Generate summary for Temperature Data
summary_temperature <- temperature_final %>%
  summarise(
    Total_Observations = n(),
    Min_Temperature = min(TemperatureCAvg, na.rm = TRUE),
    Max_Temperature = max(TemperatureCAvg, na.rm = TRUE),
    Mean_Temperature = mean(TemperatureCAvg, na.rm = TRUE),
    Median_Temperature = median(TemperatureCAvg, na.rm = TRUE)
  )

# Generate summary for Crime Data
summary_crime <- crimes_final %>%
  summarise(
    Total_Observations = n(),
    Unique_Crime_Types = n_distinct(category),
    First_Date = min(date),
    Last_Date = max(date)
  )

# Print the summaries in a nice table format using knitr
kable(summary_temperature, caption = "Summary Statistics for Temperature Data")
kable(summary_crime, caption = "Summary Statistics for Crime Data")







```
TABLE: 

In 2023, Colchester experienced different types of weather and had to deal 
with various crime issues. Every day, the temperature was carefully noted,
ranging from a freezing -2.6°C to a warm 23.1°C, with an average of about 
10.92°C throughout the year. This detailed weather record sets the stage 
for studying connections between weather and city life, including crime.

At the same time, records of crimes were kept, totaling 6,878 incidents 
spread across 14 different types. These records cover the period from 
January 1st to December 1st, giving insights into when and where crimes 
happened. Understanding these patterns is important for predicting future
crime trends and using resources wisely. This report will look at how
changes in temperature throughout the year might affect crime rates and 
suggest ways to improve safety and city planning in Colchester.

By analyzing these records, this study aims to reveal how weather and crime
are linked, providing useful insights for local authorities and community 
leaders to address the city's challenges more effectively.


```{r, eval=TRUE}
#HISTOGRAM

# Calculate summary statistics for some relevant variables and print
weather_stats <- summary(temperature_final
      [, c("TemperatureCAvg", "Precmm", "WindkmhInt" )])
kable(weather_stats, format = "html", 
      caption = "Summary Statistics for Weather")

# Histogram plot for the average temperature
hist(temperature_final$TemperatureCAvg,
     breaks = seq(-10, 40, by = 1),
     xlim = c(-5, 30),
     ylim = c(0, 0.1),
     ylab = "Density",
     xlab = "Temperature Average (Celsius)",
     main = "Probability Density of Temperature Distribution",
     prob = TRUE,
     col = 'yellow')

```
**Temperature in Colchester:**

In Colchester, the weather in 2023 had two main patterns. It was often about 
10 degrees Celsius – not too hot or cold. This mild weather might be why people 
choose to do outdoor activities. Then there were fewer days, but still quite 
a few, when it was around 20 degrees Celsius – perfect for enjoying the sunshine.
But days with an average temperature around 25 degrees were rare, so really hot
days were special. Looking at these numbers, we can guess when people might use
more energy to heat their homes or when ice cream might sell the most!


```{r, eval=TRUE}
#HISTOGRAM

# Convert the date column to a Date object if not already
crimes_final$date <- as.Date(crimes_final$date)

# Aggregate the number of crimes per day
daily_crimes <- crimes_final %>%
  group_by(date) %>%
  summarise(daily_crime_count = n(), .groups = 'drop')

# Ensure the summary is correctly calculated
print(head(daily_crimes))
# Histogram for daily crime counts, styled to match the temperature histogram
hist(daily_crimes$daily_crime_count,
     breaks = seq(min(daily_crimes$daily_crime_count),
                  max(daily_crimes$daily_crime_count), by = 1),
     main = "Probability Density of Daily Crime Distribution",
     xlab = "Daily Crime Count",
     ylab = "Density",
     col = "grey", 
     # Use grey color for the bars to match the temperature histogram
     border = "darkblue", 
     # Use black color for the border to match the temperature histogram
     xlim = c(0, max(daily_crimes$daily_crime_count) + 1),
     # Extend x-axis a bit beyond the max count
     prob = TRUE) # Use TRUE to show density

# Since the ylim is not specified, it will be automatically 
#determined based on the data
# Optionally, if you want to add a density line, it should be black 
#to match the histogram style
lines(density(daily_crimes$daily_crime_count, na.rm = TRUE), 
      col = "red", lwd = 2)
```
**Crime in Colchester:**

In Colchester, while looking at the crime reports for 2023, we notice that most 
days were quite peaceful, with crime counts typically below 100. This could mean
that for a good part of the year, the community enjoyed a calm environment. 
However, there were a few exceptions. On some rare days, the crime count shot up
to more than 500. These unusual spikes could have been due to special events or
holidays when more people were out and about, which might have led to more 
chances for crimes to happen. Understanding these patterns can help the police
prepare better for these busy days.

```{r, eval=TRUE}
#CORRELATION

var_weather <- temperature_final[, c("TemperatureCAvg","TdAvgC", "HrAvg",
                                     "PresslevHp","Precmm","VisKm","WindkmhInt", 
                                     "TemperatureCMax", "TemperatureCMin",
                                     "TotClOct", "SunD1h")]
# Compute the correlation matrix
cor_matrix <- round(cor(var_weather), 1)

# Adjust the size of the labels to prevent overlap, if needed
label_size <- 4  # Adjust this number as needed to prevent label overlap

# Visualize the correlation matrix
G4 = ggcorrplot(cor_matrix, hc.order = TRUE, type = "lower", lab = TRUE,
                lab_size = label_size, # Use the label size defined above
                outline.col = "black", ggtheme = ggplot2::theme_bw(),
                colors = c("#2166ac", "white", "#67a9cf")) # Arranged colors to a blue gradient

# Add a title to the plot
G4 <- G4 + ggtitle("Correlation Analysis between Weather Variables")

# Only use ggplotly if HTML output is expected
if (knitr::is_html_output()) {
  ggplotly(G4)
} else {
  G4
}



```
**Correlation :**

Temperature Variables: There's a strong positive correlation (0.9 and above) 
between TemperatureCMax, TemperatureCAvg, and TemperatureCMin. This is expected 
as they are measures of temperature at different times of the day.

Humidity and Dew Point: HrAvg (average humidity) shows a strong positive
correlation with TdAvgC (average temperature of dew point), which is logical 
since humidity is related to the presence of moisture (dew) in the air.

Humidity and Precipitation: HrAvg is moderately correlated with Precmm
(precipitation in millimeters). This suggests that higher humidity might 
be associated with more rainfall, which is consistent with what we know 
about weather patterns.

Visibility and Humidity: VisKm (visibility in kilometers) has a negative 
correlation with HrAvg, indicating that higher humidity may be associated with
lower visibility, likely due to fog or mist.

Cloud Cover: TotClOct (total cloud cover in eighths of the sky) does not have a 
strong correlation with most other variables, except for a moderate negative
correlation with SunD1h (sunshine duration in hours). More cloud cover naturally
leads to less sunshine.

Temperature and Pressure: There is a small positive correlation between 
temperature (TemperatureCAvg) and pressure (PresslevHp), 
but it is not very strong.



```{r, eval=TRUE}
#PIE CHART

# Convert the 'category' column to a factor if it's not already.
crimes_final$category <- factor(crimes_final$category)

# Calculate the percentage for each category.
category_perc <- round(100 * prop.table(table(crimes_final$category)), 2)

# Order the categories by percentage in decreasing order.
ordered_perc <- category_perc[order(category_perc, decreasing = TRUE)]

# Create a data frame with category names and percentages.
category_data <- data.frame(
  Category = names(ordered_perc),
  Percentage = as.numeric(ordered_perc)
)

# Use a distinct color palette - RColorBrewer can help.
num_categories <- length(unique(category_data$Category))
colors <- RColorBrewer::brewer.pal(min(num_categories, 12), "Set3")

# If you have more than 12 categories, we will need to repeat the palette.
if (num_categories > 12) {
  colors <- colorRampPalette(colors)(num_categories)
}

# Create the static pie chart using ggplot2
pie_chart <- ggplot(category_data, aes(x = "", y = Percentage, fill = Category)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start = 0) +
  theme_void() +
  scale_fill_manual(values = colors) +
  theme(legend.position = "bottom") +
  labs(fill = "Category", title = 'Percentage of Crimes by Category')

# Display the static pie chart.
pie_chart

```
**Crime Patterns in Colchester: A Reflection of Social Dynamics**

As the sun rises over Colchester, it sheds light not only on the city's
historical paths but also on its present-day streets—a scene where each
piece of crime data contributes to the city's story. The pie chart in 
front of us, filled with different colors, tells stories of challenges and
successes, struggles and hard work, all shown through the numbers it presents.

The biggest part of our pie chart is taken up by violent crime, making up 38.3%.
It shows a town dealing with aggression, where every number represents a call
for understanding, action, and solutions that bring people together and heal
wounds. This high level of violence calls for strong community efforts—programs 
that not only patrol the streets but also offer support to homes, providing
comfort and strength to those affected.

Following closely are issues like anti-social behavior and criminal damage,
including arson, covering nearly a fifth of our chart. These are signs of
unrest, voices that are not heard or addressed, perhaps stemming from changes 
in society, economic uncertainty, or the struggles of young people trying to
find their way. These issues need intervention, creating spaces for engagement
and opportunities for creativity instead of chaos.

Shoplifting, accounting for 8.05%, indicates silent struggles within markets 
and malls, suggesting unmet needs or inequalities in the digital age. It prompts
us to take a closer look at our economy and social safety nets.

And let's not forget public order offenses, at 7.73%, which reflect disturbances
spilling over into public spaces, showing a community coping with the stresses
of societal changes, possibly still adjusting to life after the pandemic.

Each percentage point, each category, is more than just a number—it's a story,
a voice, a life intersecting with the law. Together, they paint a picture of
community life, urging us to take action. We need to create policies that
protect, develop programs that empower, and build a network of support and
unity. We should walk the streets of Colchester not in fear but with 
determination to build a town based on safety and peace.

As we plan for the future, let the pie chart not just be a record of the past
but a guide toward what can be—a Colchester united in its diversity, secure
in its neighborhoods, and peaceful under its skies.

```{r, eval=TRUE}
#TIME SERIES 

# Compute the number of crimes per month by grouping some categories
crime_data_group2 <- crimes_final %>%
  group_by(date, category) %>%
  summarise(num_of_cat_crimes = n(), .groups = "drop")

# Select the category to plot and compare
selected_categories <- c("violent-crime", "anti-social-behaviour")

# Filter the data for the selected categories
selected_categories_data <- crime_data_group2 %>%
  filter(category %in% selected_categories)

# Combine original data into an xts object
tsnums3 <- xts(data.frame(num_of_cat_crimes = 
                            selected_categories_data$num_of_cat_crimes),
                            order.by = as.Date(selected_categories_data$date))

ts_df3 <- as.data.frame(tsnums3) # change to xts to data frame again

ts_df3$category <- selected_categories_data$category

# Plot using ggplot2 only
tsplot3 <- ggplot(ts_df3, aes(x = index(tsnums3),
                              y = num_of_cat_crimes, color = category)) +
                              geom_line(linewidth = 0.8) +
             geom_smooth(method = "loess", se = FALSE, aes(group = category)) +
              # Add LOESS smoothing to both categories
  labs(x = "Date", y = "Number of Crimes", 
       title = "Number of Category Crimes per Month/2023",
       color = "Category") +
  theme_bw() +
  theme(legend.position = "right", plot.title = element_text(hjust = 0.5))

# Print the static plot
tsplot3


```
**Parallel Patterns in Crime Trends: Violent Acts and Social Disruptions**

The chart before us lays out the ups and downs of two types of crimes
in Colchester during 2023: violent crimes and anti-social behavior. 
The lines on the graph, one for each crime type, seem to move together
through the months. Violent crimes happen more often, but both kinds
seem to get higher or lower at the same times.

At the start of the year, we see fewer instances of anti-social behavior
while violent crimes don't drop much. This difference tells us that what 
causes these two might not be the same. There's a noticeable increase in
violent crimes around the middle of the year — a time when people are 
likely out and about more because of the warmer weather.

This pattern can give the police and community leaders important clues.
It seems there might be specific times when it's a good idea to have more 
activities and support available to help keep the peace. The aim is to 
find out what causes these crimes to happen at the same time and figure
out how to prevent them.

By looking at this chart, we can better understand when and why crimes
might happen in Colchester. This can help us make plans to keep our
streets safer throughout the year.






```{r, eval=TRUE}
#SCATTER PLOT

# Group and summarize the temperature data by month
monthly_temperature <- temperature_final %>%
  group_by(Month = floor_date(Date, "month")) %>%
  summarise(AverageTemp = mean(TemperatureCAvg, na.rm = TRUE), .groups = "drop")

# Group and summarize the crime data by month
monthly_crime_totals <- crimes_final %>%
  group_by(Month = floor_date(date, "month")) %>%
  summarise(total_crimes = n(), .groups = "drop")

# Merge the temperature and crime data by month
combined_data <- merge(monthly_temperature, monthly_crime_totals, by = "Month")

# Create a scatter plot to compare temperatures and crime rates
scatter_plot <- ggplot(combined_data, aes(x = AverageTemp, y = total_crimes)) +
                geom_point(color = "blue", alpha = 0.7) +
             labs(x = "Average Temperature (°C)", y = "Total Number of Crimes",
          title = "Comparison of Average Temperature and Total Crime Counts") +
          theme_minimal()

# Print the scatter plot
print(scatter_plot)

# Calculate the correlation coefficient
correlation <- cor(combined_data$AverageTemp, combined_data$total_crimes)

# Print the correlation coefficient
print(correlation) #0.2420076





```



**Exploring the Warmth of Crime: **
**The Subtle Influence of Temperature on Crime Rates**

Our look at Colchester's crime data for 2023 shows something curious: when
it gets warmer, it looks like crime goes up a bit too. The graph we've made
has a bunch of dots that are supposed to tell us if there's a link between
the average temperature and how many crimes happen. There's a slight trend
that on hotter days, there might be a few more crimes than usual.

But, here’s the thing—this link is really weak. There's a correlation
coefficient of 0.24, indicating a weak relationship. The trend isn't
strong, and the dots are pretty spread out. That means we can't just
say it's the temperature causing more crimes. There's a whole bunch of 
other stuff that could be playing a part, like what's going on in the
community, whether people have jobs, and how well the police are doing their job.

So, what's our takeaway from this? Well, if we want to keep our streets safer,
we shouldn't just look at the weather report. We need to think about all the
different things that could be leading to trouble and deal with those too.
And we need to keep an eye on this as the year goes by, to make sure we're
doing all we can to nip crime in the bud, come rain or shine.
```{r, eval=TRUE}
#BOX PLOT

# Aggregate crime data to get total crimes per day
daily_crime_counts <- crimes_final %>%
  group_by(date) %>%
  summarise(TotalCrimes = n(), .groups = "drop")

# Establish a threshold for high crime days using the median
crime_threshold <- median(daily_crime_counts$TotalCrimes, na.rm = TRUE)

# Categorize days as High or Low crime
daily_crime_counts <- daily_crime_counts %>%
  mutate(CrimeLevel = ifelse(TotalCrimes > crime_threshold, 
                             "High Crime", "Low Crime"))

# Merge the temperature data with the crime level data
# Make sure both datasets have the date formatted identically 
# and the columns properly named
combined_data <- temperature_final %>%
  rename(date = Date) %>%  # Make sure Date column is the same in both datasets
  inner_join(daily_crime_counts, by = "date")

# Create the boxplot using the correct temperature column name
boxplot <- ggplot(combined_data, aes(x = CrimeLevel, 
                                     y = TemperatureCAvg, fill = CrimeLevel)) +
                geom_boxplot() +labs(x = "Crime Level", y = "Temperature (°C)",
               title = "Temperature Distribution on High vs. Low Crime Days") +
               theme_minimal() + scale_fill_brewer(palette = "Set1")

# Print the plot
print(boxplot)


```
**Temperature Tendencies on Divergent Crime Days**

Our analysis of Colchester’s 2023 crime data reveals a notable trend: days with
higher crime rates tend to be warmer. The red box, representing high crime days,
shows a temperature distribution that's generally higher compared to low crime
days, depicted in blue.

This suggests that warmer temperatures might have a role in driving crime,
possibly because more people are outdoors or there's increased social 
interaction. However, it's important to note that there's overlap in 
temperature ranges between high and low crime days, cautioning against
drawing definitive conclusions.

In light of this trend, policing strategies could consider increasing vigilance
on warmer days when temperatures exceed the median range. Nevertheless, it's 
crucial to interpret these findings with caution, as they indicate a 
correlation, not causation, and multiple factors contribute to crime rates.



```{r, eval=TRUE}
#DENSITY PLOT

# Merge temperature and crime datasets by date
merged_data <- merge(temperature_final, crimes_final, 
                     by.x = "Date", by.y = "date", all = TRUE)

# Plot density plot
density_plot <- ggplot(merged_data, aes(x = TotClOct, y = ..density..)) +
  geom_density(fill = "lightblue", alpha = 0.6) +
  labs(x = "Total Cloud Cover (%)", y = "Density",
       title = "Density Plot of Crime Rates by Total Cloud Cover") +
  theme_minimal()

# Print the plot
print(density_plot)





```
**Clouds Over Colchester: Do They Cast a Shadow on Crime?**

Looking at Colchester's crime data for the year, there seems to be a connection
between crime rates and different levels of cloud cover. The density plot shows
that there are certain points where more crimes tend to happen, depending on the
amount of cloud cover.

Rather than a straightforward relationship where more clouds mean more crime, 
the plot suggests that there are specific levels of cloud cover that coincide 
with higher crime rates. This could mean that certain weather conditions, like
when it's not completely clear but not fully cloudy either, might be linked 
to more criminal activity.

It's important to remember that just because there's a pattern in the plot, it 
doesn't mean that cloud cover directly causes crime. There could be other
factors involved. For example, maybe certain weather conditions make people 
more likely to commit crimes, or they affect how likely someone 
is to become a victim.

This initial look at the data raises questions and shows that we need to 
dig deeper to understand how weather, along with other social and environmental 
factors, might influence crime in Colchester.


```{r, eval=TRUE}
#PAIR PLOT

# Load RColorBrewer for color palettes
library(RColorBrewer)
palette_name <- "Paired"

# Generate the color palette
palette_colors <- brewer.pal(n = 11, name = palette_name)

# Selecting only the weather variables for the scatter plot 
#matrix from the final cleaned dataset
var_weather <- temperature_final[, c("TemperatureCAvg", "TdAvgC", "HrAvg",
                                     "PresslevHp","Precmm", "VisKm", 
                                      "WindkmhInt", "TemperatureCMax", 
                                     "TemperatureCMin", "TotClOct", "SunD1h")]

# Create the scatter plot matrix with colors
pairs(var_weather, col = palette_colors)



```
**Decoding Colchester's Weather Patterns**

In our climate study, we've uncovered a fascinating scatter plot matrix,
each color representing a different weather pairing. This intricate grid 
helps us understand the relationships between temperature, humidity,
pressure, and more.

For example, the connection between maximum and minimum temperatures is 
clear—a constant from morning to night. Similarly, the interplay of dew
point temperature and humidity mirrors our misty mornings and dewy 
landscapes.

The scatter plots also reveal how pressure levels relate to precipitation, 
offering insights into sudden rain or dry spells.

Each dot on these plots marks a day in Colchester's life, distinguished by 
color to show different relationships. This isn't just data—it's the heartbeat
of our town, the changing seasons, and the whispers of the wind,
captured in charts.

Our scatter plot matrix isn't just an image; it's a conversation starter—a
puzzle waiting to be solved. It prompts questions about correlations and how 
weather patterns affect daily life.

Ultimately, this visual representation of Colchester's climate is a guide to 
our future, helping us navigate the ever-changing atmosphere.




```{r, eval=TRUE}
 #SCATTER PLOT

# Group the crime data by street name and calculate the total number of crimes for each
crime_by_street <- crimes_final %>%
  group_by(street_id, street_name) %>%
  summarise(TotalCrimes = n())

# Sort the data by total number of crimes in descending order
sorted_streets <- crime_by_street %>%
  arrange(desc(TotalCrimes))

# Select the top 5 streets with the highest crime rates
top5_streets <- head(sorted_streets, n = 5)

# Create a scatter plot to visualize the total number of crimes by street name for the top 5 streets
scatterplot <- ggplot(top5_streets, aes(x = street_name, y = TotalCrimes, label = street_id)) +
  geom_point(color = "darkred", size = 3) +
  geom_text(size = 3, vjust = -0.5, hjust = 1) +  # Add street IDs as labels
  labs(x = "Street Name", y = "Total Number of Crimes",
       title = "Top 5 Streets with Highest Crime Rates") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate x-axis labels for better readability

# Print the scatter plot
print(scatterplot)


```
**Whispers from Colchester's Shadows: Uncovering Stories of the Streets**

In the heart of Colchester, where history meets modern life, scattered points
on our map reveal a truth often overlooked, marked in cautionary red. These 
points, like signals, shed light on stories from Blackene Gardens, Church 
Street, Cowdray Avenue, the busy Shopping Area, and St Nicholas Street—each 
a part of the city's complex crime story.

In Blackene Gardens, the red point reflects tales of hidden alleys and quick
thefts disappearing like shadows at dawn. Church Street, with its history and
lively atmosphere, has seen tensions rise occasionally leading to conflicts.

As night falls, Cowdray Avenue changes. Here, the red point marks chance
encounters and secretive dealings. Similarly, the busy Shopping Area has 
hidden spots where thefts occur unnoticed.Perhaps more precautions should 
be taken for these streets with the highest crime rates.

```{r, eval=TRUE}
#LEAFLET

library(leaflet)

# Initialize the map with base tiles
map <- leaflet(crimes_final) %>% 
  addTiles(group = "OSM (default)") %>%
  addProviderTiles(providers$Stadia.AlidadeSmooth, group = "Alidade Smooth") %>%
  setView(lng = 0.901230, lat = 51.889801, zoom = 13)

# Add circles on the map for each crime category with vibrant colors
map1 <- map %>%
addCircles(data = crimes_final[crimes_final$category == "bicycle-theft",],
             ~long, ~lat, group = "Bicycle Theft", col = "#e63946",
             popup = ~paste("Bicycle Theft", "<br>Street Name:",
                            street_name)) %>%
addCircles(data = crimes_final[crimes_final$category == "drugs",], 
             ~long, ~lat, group = "Drugs", col = "#f4a261",
             popup = ~paste("Drugs", "<br>Street Name:", street_name)) %>%
addCircles(data = crimes_final[crimes_final$category == "anti-social-behaviour",], 
             ~long, ~lat, group = "Anti-Social Behaviour", col = "#2a9d8f",
             popup = ~paste("Anti-Social Behaviour", "<br>Street Name:",
                            street_name)) %>%
addCircles(data = crimes_final[crimes_final$category == "robbery",], 
             ~long, ~lat, group = "Robbery", col = "#264653",
             popup = ~paste("Robbery", "<br>Street Name:", street_name)) %>%
addCircles(data = crimes_final[crimes_final$category == "vehicle-crime",], 
             ~long, ~lat, group = "Vehicle Crime", col = "#2b2d42",
             popup = ~paste("Vehicle Crime", "<br>Street Name:", 
                            street_name)) %>%
addCircles(data = crimes_final[crimes_final$category == "theft-from-the-person",], 
             ~long, ~lat, group = "Theft from the Person", col = "#8d99ae",
             popup = ~paste("Theft from the Person", "<br>Street Name:",
                            street_name)) %>%
addCircles(data = crimes_final[crimes_final$category == "criminal-damage-arson",], 
             ~long, ~lat, group = "Criminal Damage", col = "#ef233c",
             popup = ~paste("Criminal Damage", "<br>Street Name:",
                            street_name)) %>%
addCircles(data = crimes_final[crimes_final$category == "burglary",], 
             ~long, ~lat, group = "Burglary", col = "#d00000",
             popup = ~paste("Burglary", "<br>Street Name:", street_name)) %>%
addCircles(data = crimes_final[crimes_final$category == "possession-of-weapons",], 
             ~long, ~lat, group = "Possession of Weapons", col = "#023e8a",
             popup = ~paste("Possession of Weapons", "<br>Street Name:",
                            street_name)) %>%
addCircles(data = crimes_final[crimes_final$category == "shoplifting",], 
             ~long, ~lat, group = "Shoplifting", col = "#0077b6",
             popup = ~paste("Shoplifting", "<br>Street Name:", street_name)) %>%
 addCircles(data = crimes_final[crimes_final$category == "violent-crime",], 
             ~long, ~lat, group = "Violent Crime", col = "#03045e",
             popup = ~paste("Violent Crime", "<br>Street Name:", street_name))

# Control layers for toggling visibility of crime categories
map1 %>%
  addLayersControl(
    baseGroups = c("OSM (default)", "Alidade Smooth"),
    overlayGroups = c("Bicycle Theft", "Drugs", "Anti-Social Behaviour", 
                      "Robbery", "Vehicle Crime", "Theft from the Person", 
                      "Criminal Damage", "Burglary", "Possession of Weapons", 
                      "Shoplifting", "Violent Crime"),
    options = layersControlOptions(collapsed = FALSE))





```


**Navigating the Crime Landscape: Insights from Colchester’s Interactive Map**

In our quest to understand the spatial dynamics of criminal activity within
Colchester, we have employed an interactive leaflet map that reveals the
geographical nuances of crime distribution. Each category of crime is
illuminated by a specific hue, turning our city map into a mosaic of insights:

**Bicycle Theft (Red, #e63946):** Marked in an unmistakable red, bicycle thefts dot
the landscape, with clusters appearing around key transit routes and
recreational areas, perhaps hinting at opportunistic crime hotspots.

**Drugs (Orange, #f4a261):** The orange circles give us a view into the locations
where drug-related incidents are most prevalent, possibly correlating with
nightlife hubs or areas of lesser vigilance.

**Anti-Social Behaviour (Teal, #2a9d8f):** Teal marks anti-social behavior, 
painting a picture of social discontent in specific neighborhoods, potentially 
guiding community outreach initiatives.

**Robbery (Dark Slate, #264653):** In dark slate, robberies are charted, often
concentrated in commercial districts, suggesting the need for enhanced security
measures in these regions.

**Vehicle Crime (Charcoal, #2b2d42):** The charcoal circles that mark 
vehicle-related crimes tend to cluster around major parking facilities 
and thoroughfares.

**Theft from the Person (Slate Gray, #8d99ae):** Represented in slate gray,
theft from the person incidents can often be found in the busier parts of the
city, where pedestrian traffic is high.

**Criminal Damage and Arson (Crimson, #ef233c):** Spots of crimson expose areas 
afflicted by property damage and arson, demanding attention to urban decay or 
the need for conflict resolution.

**Burglary (Deep Red, #d00000):** Deep red circles indicate burglary, which may 
align with residential areas, bringing to light the vulnerabilities
of home security.

**Possession of Weapons (Navy, #023e8a):** The navy color used for possession of
weapons can be a marker for areas that might benefit from preventative policing
and educational programs.

**Shoplifting (Blue, #0077b6):** The blue points illustrate shoplifting occurrences,
predictably found in retail-heavy zones, signaling loss prevention as a priority.

**Violent Crime (Dark Blue, #03045e):** Lastly, violent crime is denoted in dark blue,
often dispersed throughout the city but with notable concentrations that could 
require a multi-faceted approach to curb


